---
title: "Bandwidths for Univariate Kernel Density Estimation"
author: "Brad Stieber"
date: ""
output: 
  beamer_presentation: 
    keep_tex: yes
---


#Introduction

\begin{columns}
\begin{column}{0.5\textwidth}
    KDE (choose kernel $K$ and bandwidth $h$): 
    $$ \widehat{f_h}(x) = n^{-1} \sum_{i = 1}^{n}K\left(\frac{x - x_i}{h}\right) $$
    Optimal Bandwidth:
    $$ h_{AMISE} = \left(\frac{R(K)}{n\sigma_K^4R(f^{''})}\right)^{\frac{1}{5}} $$
    
    - $R(g) = \int g^2$: roughness of a function
    
    - Don't know $R(f^{''})$ $\rightarrow$ bandwidth selections rely on getting around this unknown
    
\end{column}
\begin{column}{0.5\textwidth}  %%<--- here
    \begin{figure}[t]
     \includegraphics[width = 7in]{KDEDemo.png}
     \centering
     \end{figure}
\end{column}
\end{columns}



# Candidate Bandwidths (1/2)

## Unbiased Cross Validation (E[UCV + R(f)] = MISE)

Minimize 

$$
UCV(h) = R\left(\widehat{f}\right) - \frac{2}{n} \sum_{i = 1}^{n} \widehat{f}_{-i}(x_i),
$$
instead.

$$
\widehat{f}_{-i}(x_i) = \frac{1}{h(n-1)}\sum_{j \neq i} K\left(\frac{x_i - x_j}{h}\right)
$$


is the LOO estimator. Used to estimate the second term in $ISE(h) = \int \widehat{f}^2 - 2 \int\widehat{f}_h f + \int f^2$.

Issue: excessive variation

# Candidate Bandwidths (2/2)

## Terrell's Maximal Smoothing

Instead of estimating $R(f^{''})$, what if we tried to minimize it? Built on the result that the $beta(k + 2, k + 2)$ family minimizes $\int (f^{(k)})^2$ for a given standard deviation.

$$
h_{MS} =  3 \hat \sigma \left(\frac{R(K)}{35 n} \right)^{\frac{1}{5}}.
$$
Issue: upper bound on $h_{opt}$ $\rightarrow$ oversmooths interesting features of the data


```{r echo = FALSE, message = FALSE, warning=FALSE}
library(ggplot2)

bw.MS <- function(x) {3 * ((2 * sqrt(pi))^(-1) / (35 * length(x)))^(1/5) * sd(x)}

#generate sample data

#example 1 (mixture of normals)

set.seed(123)
x_1 <- c(rnorm(150), rnorm(150, mean = 4, sd = 2))
df_x <- data.frame(x = x_1)
n <- length(x_1)

from_1 <- min(x_1) - .5 * sd(x_1)
to_1 <- max(x_1) + .5 * sd(x_1)

h.MS <- bw.MS(x_1)
h.UCV <- bw.ucv(x_1)

dens.MS <- density(x_1, from = from_1, to = to_1,
                   bw = h.MS, n = 1000)

dens.UCV <- density(x_1, from = from_1, to = to_1,
                    bw = h.UCV, n = 1000)

#create data.frame

df_dens <- data.frame(h = rep(
    c(paste('MS:', round(h.MS, 2)),
      paste('UCV:', round(h.UCV, 2))),
    each = length(dens.MS$x)),
    x = c(dens.MS$x,
          dens.UCV$x),
    y = c(dens.MS$y, dens.UCV$y),
    stringsAsFactors = FALSE)

#make plot

p1 <- ggplot()+
    geom_histogram(data = df_x, aes(x = x, y = ..density..),
                   bins = 1.4 * ceiling(sqrt(n)),
                   fill = NA, colour = 'black')+
    geom_line(data = df_dens, aes(x = x, y = y),
              size = 1.2, colour = 'dodgerblue4')+
    theme_bw()+
    theme(panel.grid = element_blank(),
          axis.title = element_text(size = 16),
          strip.text = element_text(size = 16, face = 'bold'),
          axis.text = element_text(size = 14))+
    xlab('X')+ylab('Density')+
    facet_wrap(~h)


#another data set
library(MASS)

x_1 <- geyser$waiting
df_x <- data.frame(x = x_1)
n <- length(x_1)

from_1 <- min(x_1) - .5 * sd(x_1)
to_1 <- max(x_1) + .5 * sd(x_1)

h.MS <- bw.MS(x_1)
h.UCV <- bw.ucv(x_1)

dens.MS <- density(x_1, from = from_1, to = to_1,
                   bw = h.MS, n = 1000)

dens.UCV <- density(x_1, from = from_1, to = to_1,
                    bw = h.UCV, n = 1000)

#create data.frame

df_dens <- data.frame(h = rep(
    c(paste('MS:', round(h.MS, 2)),
      paste('UCV:', round(h.UCV, 2))),
    each = length(dens.MS$x)),
    x = c(dens.MS$x,
          dens.UCV$x),
    y = c(dens.MS$y, dens.UCV$y),
    stringsAsFactors = FALSE)

#make plot

p2 <- ggplot()+
    geom_histogram(data = df_x, aes(x = x, y = ..density..),
                   bins = 1.4 * ceiling(sqrt(n)),
                   fill = NA, colour = 'black')+
    geom_line(data = df_dens, aes(x = x, y = y),
              size = 1.2, colour = 'dodgerblue4')+
    theme_bw()+
    theme(panel.grid = element_blank(),
          axis.title = element_text(size = 16),
          strip.text = element_text(size = 16, face = 'bold'),
          axis.text = element_text(size = 14))+
    xlab('Waiting Time')+ylab('Density')+
    facet_wrap(~h)


#another data set
library(RCurl)
git_file <- 
    getURL("https://raw.githubusercontent.com/bgstieber/Top250Beer/master/Top250Beers.csv")
beer_data <- read.csv(text = git_file, stringsAsFactors = FALSE)
x_1 <- beer_data[!is.na(beer_data$ABV), ]$ABV

df_x <- data.frame(x = x_1)
n <- length(x_1)

from_1 <- min(x_1) - .1 * sd(x_1)
to_1 <- max(x_1) + .1 * sd(x_1)

h.MS <- bw.MS(x_1)
h.UCV <- bw.ucv(x_1)

dens.MS <- density(x_1, from = from_1, to = to_1,
                   bw = h.MS, n = 1000)

dens.UCV <- density(x_1, from = from_1, to = to_1,
                    bw = h.UCV, n = 1000)

#create data.frame

df_dens <- data.frame(h = rep(
    c(paste('MS:', round(h.MS, 2)),
      paste('UCV:', round(h.UCV, 2))),
    each = length(dens.MS$x)),
    x = c(dens.MS$x,
          dens.UCV$x),
    y = c(dens.MS$y, dens.UCV$y),
    stringsAsFactors = FALSE)

#make plot

p3 <- ggplot()+
    geom_histogram(data = df_x, aes(x = x, y = ..density..),
                   bins = 1.4 * ceiling(sqrt(n)),
                   fill = NA, colour = 'black')+
    geom_line(data = df_dens, aes(x = x, y = y),
              size = 1.2, colour = 'dodgerblue4')+
    theme_bw()+
    theme(panel.grid = element_blank(),
          axis.title = element_text(size = 16),
          strip.text = element_text(size = 16, face = 'bold'),
          axis.text = element_text(size = 14))+
    xlab('Alcohol by Volume (%)')+ylab('Density')+
    facet_wrap(~h)

```


# Example 1

50/50 mixture of $N(0,1)$ and $N(2, 2^2)$

```{r echo = FALSE}
p1
```

<!--
# Example 2

Waiting times for eruptions at Old Faithful geyser in WY (1st through 15th of Aug. 1985)

```{r echo = FALSE}
p2
```


-->

# Example 2

Alcohol by volume of Beer Advocate's top 250 beers

```{r echo = FALSE}
p3
```


# Conclusion

- Choosing a bandwidth should be an iterative process
- Bias - variance tradeoff
    - Too smooth: low variance, high bias
        - what _is not there_ might be
    - Too wiggly: high variance, low bias
        - what _is there_ might be too hard to see
        
        
        
